Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-30 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical inferences
derived from these processes.

@@

Thurstone noted from the outset that his general form, described in
Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ experiment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations --one per comparison (see Table~\ref{tbl-thurstone_cases}).
However, a single judge could only provide \({5 \choose 2} = 10\) unique
comparisons, an insufficient data set to estimate the required
parameters.

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge with its
measurement model: it required estimating more ``unknown'' parameters
than the number of available pairwise comparisons
\citep{Thurstone_1927b}. To address this issue and facilitate the
theory's practical applicability, he formulated five cases, each
progressively incorporating several simplifying assumptions.

Among these, Case V has become the most widely used model in empirical
CJ research, primarily due to the widespread adoption of the BTL model.
The BTL model mirrors the core assumptions of Case V--namely, equal
discriminal dispersions and zero correlation among stimuli' discriminal
processes--but replaces the processes' normal distribution with the more
mathematically tractable logistic distribution
\citep{Andrich_1978, Bramley_2008}. Although the substitution has a
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021} and simplifies the
measurement model, the other two core assumptions of the model can still
lead to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023},
especially when researchers assess complex traits or heterogeneous
stimuli
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022}.

Furthermore, because Thurstone primarily aimed to produce a trait's
``coarse scaling'' and allocate the compared stimuli along this
continuum \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the trait estimates for statistical inference. The CJ
tradition has attempted to fill this gap by separating trait estimation
from hypothesis testing. Specifically, researchers often rely on point
estimates of the traits, such as the BTL scores or their
transformations, for inference. Although this approach simplifies
analysis, the statistical literature cautions that it can introduce bias
and compromise the reliability of the resulting inferences
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The study then translates the SCM into a bespoke statistical model that
addresses key limitations of Case V. This model allows researchers to
analyze CJ data when the assumptions of equal dispersion and zero
correlation do not hold (see Section~\ref{sec-theory-issue1a} and
Section~\ref{sec-theory-issue1b}), and when they need to perform
statistical inference (refer to Section~\ref{sec-theory-issue2}). In
particular, the model accounts for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), captures the hierarchical
structure of stimuli and incorporates measurement error in the
hypothesis testing process (refer to
Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
in discriminal dispersions (see Section~\ref{sec-statistical}). These
methodological innovations can potentially enhance the reliability and
validity of trait measurement \citep{Perron_et_al_2015} and improve the
accuracy of statistical inferences.

@@

Sampling and comparison mechanisms play a central role in modern CJ
experimental design. However, most of the CJ literature has examined
these mechanisms within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}.
Although these studies offer valuable insights, we argue that they may
have overlooked the broader role these mechanisms play within the CJ
system. Moreover, we suggest that this oversight largely stems from a
lack of conceptual clarity about how the mechanisms function within the
system. To address this gap, this study decided to integrate these
mechanisms into CJ's conceptual model.

@@

However, many modern CJ applications rely on more complex experimental
designs in which the sampling and comparison mechanisms introduce more
intricate forms of missingness--such as \emph{missing at random} (MAR)
or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
prominent example is the previously discussed ACJ designs, where prior
judgment outcomes inform the selection of stimulus pairs for subsequent
comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
outcome-informed selection suggests that the comparison mechanism may be
outcome-dependent, potentially classifying ACJ as a generator of MNAR
data. This classification may, in turn, facilitate a more nuanced
understanding of the mixed evidence surrounding ACJ's capabilities.
While some studies report improvements in trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
that these gains are artificially inflated by the method
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.

Nevertheless, regardless of the underlying missingness mechanisms, any
CJ experimental design would benefit from explicitly defining its
assumptions--an approach supported by our framework. This clarity
enables researchers to evaluate, for instance, how sampling and
comparison mechanisms affect trait estimation and statistical inference
within each design. Such assessments are particularly relevant given the
common misconception in the CJ literature that Thurstone's model can
naturally handle even non-random missing data without compromising the
reliability or validity of trait estimates \citep{Bramley_2008}.

@@

Despite the growing notion that various stimulus-related factors can
influence judges' perceptions during pairwise comparisons--and that
these influences may not always cancel each other out--few studies in
the CJ literature provide empirical evidence for judges' biases
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
We argue that this gap persists not due to a lack of interest or
research but because researchers continue to rely on ad-hoc detection
methods, such as `misfit' statistics, that are poorly suited for the
task \citep{Kelly_et_al_2022}. To address this limitation, this study
decided to treat judges' biases as an integral component of the CJ
system from the outset. This approach offers one key advantage: it
provides a more accurate representation of the data-generating process
behind pairwise comparisons--one that acknowledges that the discriminal
processes of stimuli become an observable outcome only through judges'
perceptions, which may exhibit bias.

@@

Moreover, if we accept that judges may value different aspects of a
stimulus, it follows that their judgments might also vary with
individual characteristics such as gender, age, culture, income,
education, training, or expertise
\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}, and even from one
individual to another
\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
Building on this and the discussion in Section~\ref{sec-discussion_RA1},
researchers could investigate how the selection of judges influences the
formation of a ``shared consensus'' and whether these individual
characteristics introduce systematic biases or distortions in the
observed trait distribution \citep{Deffner_et_al_2022}. Furthermore, if
such characteristics indeed undermine the ``sample-freeness''
assumption, then it becomes essential to explore strategies for
controlling their influence. This includes investigating how many
judges--and how many judgments per judge--are required to produce
reliable trait estimates, as well as whether \emph{repeated measures
designs}, in which judges evaluate the same stimulus pairs multiple
times \citep{Lawson_2015}, can improve judgment consistency and
accuracy.

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates. First, it critiques
the heavy reliance on Thurstone's Case V assumptions
\citep{Thurstone_1927b} and, by extension, the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement model, they may fail to
capture the complexity of some traits or account for heterogeneous
stimuli, potentially leading to unreliable and inaccurate trait
estimates. Second, the study highlights the disconnect between CJ's
trait measurement and hypothesis testing processes. Although separating
these processes simplifies the analysis of CJ data, it may also
undermine the reliability of various statistical inferences derived from
these processes.

@@

Thurstone noted from the outset that his general formulation, described
in Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. For instance, in a CJ experiment with five
texts, the general form would require estimating \(20\) parameters: five
modal discriminal processes, five discriminal dispersions, and \(10\)
correlations--one per comparison (see Table~\ref{tbl-thurstone_cases}).
However, a single judge could only provide \({5 \choose 2} = 10\) unique
comparisons, an insufficient data set to estimate the required
parameters.

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge in this general
formulation: the measurement model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep{Thurstone_1927b}. To address this issue and facilitate the
theory's practical applicability, he formulated five cases, each
progressively incorporating several simplifying assumptions.

Among these, Case V remains the most widely used model in empirical CJ
research, mainly due to the widespread adoption of the BTL model. The
BTL model mirrors the core assumptions of Case V--namely, equal
discriminal dispersions and zero correlation among stimuli' discriminal
processes--but replaces the processes' normal distribution with the more
mathematically tractable logistic distribution
\citep{Andrich_1978, Bramley_2008}. Notably, although these assumptions
simplify the trait measurement model, they may fail to capture the
complexity of some traits or account for heterogeneous stimuli
\citep{Thurstone_1927a, Andrich_1978, Bramley_2008, Kelly_et_al_2022},
potentially leading to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Moreover, because Thurstone's original goal was to produce a ``coarse
scaling'' of traits and allocate stimuli along this continuum
\citep[pp.~269]{Thurstone_1927a}, his theory offered no guidance on how
to use trait estimates for statistical inference. To address this gap,
researchers in the CJ tradition have often separated trait estimation
from hypothesis testing, relying on point estimates of the traits--such
as BTL scores or their transformations--for inference. Notably, while
this approach simplifies analysis, it can also introduce bias and
compromise the reliability of the resulting inferences
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The study then translates the SCM into a bespoke statistical model that
allows researchers to analyze CJ data when violations to the assumptions
of equal dispersion and zero correlation occur (see
Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b})
and when statistical inference is necessary (refer to
Section~\ref{sec-theory-issue2}). In particular, this model accounts for
judge biases (see Section~\ref{sec-theory-theoretical_P1}), captures the
hierarchical structure of stimuli, incorporates measurement error into
the hypothesis testing process (refer to
Section~\ref{sec-theory-theoretical_P3}), and accommodates heterogeneity
in discriminal dispersions (see Section~\ref{sec-statistical}). By
addressing all these issues, these methodological innovations have the
potential to enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
inferences.

@@

Although sampling and comparison mechanisms are central to modern CJ
experimental design, it is striking that most of the CJ literature has
examined them only within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}.
Although these studies offer valuable insights, we argue that they have
overlooked the broader role these mechanisms play within the CJ system.
Moreover, we suggest that this oversight largely stems from a lack of
conceptual clarity about how the mechanisms function within the system.
To address this gap, this study decided to integrate these mechanisms
into CJ's conceptual model.

@@

However, many modern CJ applications rely on more complex experimental
designs in which the sampling and comparison mechanisms introduce more
intricate forms of missingness--such as \emph{missing at random} (MAR)
or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
prominent example is the previously discussed ACJ designs, where prior
judgment outcomes inform the selection of stimulus pairs for subsequent
comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
outcome-informed selection suggests that the comparison mechanism may be
outcome-dependent, potentially classifying ACJ as a generator of MNAR
data. This classification may, in turn, facilitate a more nuanced
understanding of the mixed evidence surrounding ACJ's capabilities.
While some studies report that the method improves trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
that it artificially inflates these gains
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.

Nevertheless, regardless of the underlying missingness mechanisms, any
CJ experimental design would benefit from explicitly defining its
assumptions, an approach supported by our framework. This clarity
enables researchers to evaluate, for instance, how sampling and
comparison mechanisms affect trait estimation and statistical inference
within each design. Such assessments are particularly relevant given the
common misconception in the CJ literature that Thurstone's model can
naturally handle even non-random missing data without compromising the
reliability or validity of trait estimates \citep{Bramley_2008}.

@@

Despite the growing notion that various stimulus-related factors can
influence judges' perceptions \citep{Morin_et_al_2018} and that these
influences may not always cancel each other out, few studies in the CJ
literature provide empirical evidence for judges' biases
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a}.
We argue that this gap persists not due to a lack of interest or
research but because researchers continue to rely on ad-hoc detection
methods, such as `misfit' statistics, that are poorly suited for the
task \citep{Kelly_et_al_2022}. To address this limitation, this study
decided to treat judges' biases as an integral component of the CJ
system from the outset. This approach offers one key advantage: it
provides a more accurate representation of the data-generating process
behind pairwise comparisons, one that acknowledges that the discriminal
processes of stimuli become an observable outcome only through judges'
perceptions, which may exhibit bias.

@@

Moreover, if we accept that judges may prioritize different aspects of a
stimulus, it follows that their assessments might also vary with
individual characteristics such as gender, age, culture, income,
education, training, or expertise
\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}, and even from one
individual to another
\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
Thus, building on the discussion in Section~\ref{sec-discussion_RA1},
researchers could investigate how the selection of judges influences the
formation of a ``shared consensus'' and whether these individual
characteristics introduce systematic biases or distortions in the
observed trait distribution \citep{Deffner_et_al_2022}. Furthermore, if
such characteristics indeed undermine the assumption of
``sample-freeness,'' then it becomes essential to explore strategies to
control their influence and to determine how many judges--and how many
judgments per judge--are needed to produce reliable trait estimates in
these contexts. Additionally, it is worth considering whether
\emph{repeated measures designs}, in which judges evaluate the same
stimulus pairs multiple times \citep{Lawson_2015}, can improve judgment
consistency and accuracy.
